Introductory statistics with `intRo`

### 2016-06-15 jcgs-paper-2016/paper.Rmd
Changes from mtg with andee

The widespread adoption of R as a tool for statistical analysis has undoubtedly been an important development for the scientific community. However, using R in most cases still requires a basic knowledge of programming concepts which may pose a steep learning curve for the introductory statistics student. This additional time commitment may explain why introductory courses often utilize software applications such as JMP, even though instructors tend to use a tool like R in their own work. Still, some compromises must be made when using software like JMP, including dealing with software licenses and unsupported desktop platforms. **AK: This is heavy on the JMP**

@@

Multiple software packages have recently been written in an attempt to spur interest in introductory statistics and R programming. DataCamp's [@datacamp] courses are user-friendly ways to learning basic R programming and data analysis techniques. Swirl [@swirl] is a similar interactive tool to make learning R more fun by learning it within R itself. Project MOSAIC [@mosaic] has created a suite of tools to simplify the teaching of statistics in the form of an R package. The primary goal of DataCamp and Swirl is to teach R programming, rather than facilitate the learning of introductory statistics. Project MOSAIC's goal is to faciliate this learning, but using the package requires a knowledge of R programming that the introductory student may not have. iNZight Lite [@inzight] also attempts to expose students to data analysis without requiring programming knowledge. While this tool has a lot of functionality and a very broad scope, it does not include reproducible R code, and therefore has less of a focus on spurring interest in coding for students **AK: this seems a little out of place, did we even mention the interest thing yet?**. R Commander [@fox2005] and Deducer [@fellows2012] provide a graphical front-end to many statistical functions in R, but are not web-based and thus require local installation and configuration.

@@

`intRo` has an ulterior motive as well: to get students excited about programming. By navigating about the user interface of `intRo`, students are actually creating a fully-executable R script that they can download and run locally as well as viewing the script change real-time within the application. This code creation element of `intRo` is meant to generate excitement about programming in R and empower students to feel that they can generate code as well. `intRo` uses rmarkdown's render function in order to print the results, by dymanically executing the student's R script. By default, the output will include the R code, but if the student elects to hide the source code by clicking the code toggle button at the top, the code will not appear in the printed results. This reproducibility framework has a number of key advantages:

1. This eases a student who may be intimidated by programming into the idea that interacting with a user interface is really just a frontend for code. Seeing the correspondence between graphical clicks and printed code will hopefully also lessen the fear of coding that many students may have.

2. An analysis created by `intRo` can be reproduced in an R session to easily assess and extend the results.

3. "Printing" the results of an `intRo` analysis amounts to nothing more than executing the R code on the server, adding another layer of reproducibility.

@@

An `intRo` module is a set of self-contained executible R scripts that together produce a set of introductory statistics functionality. `intRo` modules were designed in this way to allow for simple dynamic creation of the user interface at run-time, as well as ease the process of converting existing analysis code to the `intRo` framework. A high-level diagram of this process is given in figure \ref{fig:app_creation_modules}. Based on the modules contained within the `modules` folder, the interface and functionality of the app is dynamically generated at runtime. **AK: We need a more detailed description of this image.**

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/app_creation_modules.pdf}
\caption{This figure depicts how the \texttt{server.R} and \texttt{ui.R} files are populated using the modular structure within \texttt{intRo}.}
\label{fig:app_creation_modules}
\end{figure*}

@@

Modularity was a design decision we focused on from the start of `intRo's` development. There are some practical benefits to thinking of related statistics and data science functionality in terms of modules: **AK: We have too many lists in this paper.**

1. Course instructors can easily enable or disable functionality depending on the exact needs of the particular class being taught. Because this is done at run-time, including new modules is as simple as downloading and placing it within `intRo's` modules folder, or removing existing modules from that folder.
2. Errors can be more easily isolated to specific components. For instance, if an error is encountered, simply disabling the module can provide a temporary workaround while the issue is identified.
3. Modularity helps to organize the different pieces of code into functionality chunks that make it easier for developers to maintain.

@@

The modules provided with `intRo` are contained in the modules folder. The top level directory in the modules folder defines the category of the module (currently `data`, `summaries`, or `inference`). Within each of these categories is a folder named according to the name of the module. This folder houses the previously defined scripts. As an example, we will walk through the contents of the `contingency` module.

**AK: Should we walk through a difference models? Since this is not inference, it is confusing**
Since the `contingency` module performs a statistical analysis, it is part of the inference category, and hence can be found in the `intRo` repository at **modules/inference/contingency**. Let's first examine *helper.R*:

```{r, eval=FALSE}
cont.table <- function(intro.data, x, y, type, digits) {  
    interpolate(
        ~ (my.tbl <- cbind(rbind(table(df$y, df$x), 
                      Total = colSums(table(df$y, df$x))), 
                Total = c(rowSums(table(df$y, df$x))[-(length(df$x))], 
                          sum(table(df$y, df$x))))),
        df = quote(intro.data),
        x = x,
        y = y,
        mydir = userdir,
        `_env` = environment(),
        file = "code_contingency.R"
    )

    # Additional formatting code suppressed for readability
    ...

    interpolate(~(format(round(my.tbl, digits = new.digits))), 
                mydir = userdir, 
                `_env` = environment(), 
                file = "code_contingency.R",
                append = TRUE)
}
```

This script is most immediately similar to standard R code. In this case, a function `cont.table` is created which, depending on the values of the parameters, ultimately returns a contingency table. One important difference from a typical R script is that each call in the script is wrapped in a function called `interpolate` [@interpolate]. `interpolate` both executes the given R code on the server, and also writes the code executed to the script window at the bottom of `intRo`.

Because all the code needed to implement a contingency table is found in the `base` and `stats` package, the `libraries.R` file is empty for the `contingency` module. `observe.R`, which defines the Shiny observers needed, is reproduced below:

```{r, eval=FALSE}
observe({
    xselect <- ifelse(checkVariable(intro.data(), input$xcont),
                      input$xcont, 
                      intro.categoricnames()[1])
    yselect <- ifelse(checkVariable(intro.data(), input$ycont), 
                      input$ycont, 
                      intro.categoricnames()[2])

    updateSelectizeInput(session, 
                         "xcont", 
                         choices = setdiff(intro.categoricnames(), 
                                           yselect), 
                         selected = xselect)
    updateSelectizeInput(session, 
                         "ycont", 
                         choices = setdiff(intro.categoricnames(), 
                                           xselect), 
                         selected = yselect)
})

observeEvent(input$store_contingency, {
    cat(paste0("\n\n", paste(readLines(
        file.path(userdir, "code_contingency.R")), 
        collapse = "\n")), 
        file = file.path(userdir, "code_All.R"), 
        append = TRUE)
})
```

Shiny observers are a class of reactive objects within the Shiny paradigm which do not return a value [@shiny]. For further discussion of reactivity, see section \ref{reactive-design-choices}. In this example, observers are created to ensure that the choices of variable for the `contingency` module are only categorical variables. This is accomplished by utilizing the global reactive `intro.categoricnames()`, which returns a character vector containing the variables in the current dataset that are categorical. Finally, there is an event observer to store code generated from the module into the overall code script upon clicking the store button. The presence of this observer code and the definition of the button in the user interface are enforced, and must be present in any `intRo` module.

The `output.R` and `reactive.R` code is very simple:

```{r, eval=FALSE}
output$conttable <- renderTable({
    return(intro.contingency())
})
```

```{r, eval=FALSE}
intro.contingency <- reactive({
    if (is.null(intro.data())) return(NULL)
    if (input$xcont == input$ycont || 
        length(categoricNames(intro.data())) < 2) return(NULL)

    my.conttable <- cont.table(intro.data(), 
                               input$xcont, input$ycont, 
                               input$conttype, input$contdigits)

    return(my.conttable)
})
```

`intro.contingency` is defined as a reactive object. Some small error checks are performed at the beginning to ensure that the data is present and available, and that two different categorical variables are selected. Assuming those conditions hold, a call to the previously defined `helper.R` function is made with the appropriate parameters, and the results are returned. The `output.R` script then simply uses Shiny's `renderTable` function to display the resulting table.

@@

```{r, eval=FALSE}
contingency_ui <- tabPanel("Contingency",
    column(4,
         wellPanel(
             selectizeInput("xcont", label = "X Variable (x)", 
                            choices = categoricNames(mpg), 
                            selected = categoricNames(mpg)[5]),

             # Additional inputs suppressed for readability
             ...

@@

             tags$button("", id = "store_contingency", type = "button", 
                         class = "btn action-button", 
                         list(icon("save"), "Store Contingency Result"), 
                         onclick = "$('#top-nav a:has(> .fa-print, 
                         .fa-code, .fa-download)').highlight();")
         )
    ),

    column(8,
         tableOutput("conttable")
    )
)
```

@@

**AK: I feel like this section could use a little more beef, but I'm not sure how**

@@

We believe **AK: we believe is weak language, consider removing** `intRo` can be a powerful and effective tool for introductory statistics education. Its modular structure allows it to be flexible enough for many different applications and curriculums. Its ease-of-use allows the student to focus her attention on the statistics task at hand, rather than struggling with software licenses and confusing interface navigation. Reproducible code generated from each analysis can be used to spark an interest in R programming in those who might otherwise not be exposed to it.

@@

Challenges do exist with regards to the wider adoption of `intRo`. For instance, we will need to monitor how well the server hosting `intRo` handles the load of dozens of students performing data analyses at once. If performance issues are encountered, the infrastructure used may need to be expanded to handle current and future load. An unknown quantity will be how feasible it is to increase adoption of `intRo` across the department **AK: not sure about "the department"**, as well as to other universities. One limitation of `intRo` is that uploading a dataset beyond about 30,000 rows tends to be slow. Even once the data is successfully uploaded, the default modules produce results more slowly than with smaller datasets. This is a limitation that should be further investigated if and when `intRo` sees wider adoption.

>>>>

The widespread adoption of R as a tool for statistical analysis has undoubtedly been an important development for the scientific community. However, using R in most cases still requires a basic knowledge of programming concepts which may pose a steep learning curve for the introductory statistics student. This additional time commitment may explain why introductory courses often utilize point-and-click applications, even though instructors tend to use a tool like R in their own work. Still, some compromises must be made when using many graphical applications, including dealing with software licenses and unsupported desktop platforms.

@@

Multiple software packages have recently been written in an attempt to spur interest in  R programming and statistics. DataCamp's [@datacamp] courses are user-friendly ways to learning basic R programming and data analysis techniques. Swirl [@swirl] is a similar interactive tool to make learning R more fun by learning it within R itself. Project MOSAIC [@mosaic] has created a suite of tools to simplify the teaching of statistics in the form of an R package. The primary goal of DataCamp and Swirl is to teach R programming, rather than facilitate the learning of introductory statistics. Project MOSAIC's goal is to faciliate this learning, but using the package requires a knowledge of R programming that the introductory student may not have. R Commander [@fox2005] and Deducer [@fellows2012] provide a graphical front-end to many statistical functions in R, but are not web-based and thus require local installation and configuration. iNZight Lite [@inzight] also attempts to expose students to data analysis without requiring programming knowledge, but does not include reproducible R code, and therefore has less of a focus on spurring interest in coding for students.

@@

`intRo` has an ulterior motive as well: to get students excited about programming. By navigating about the user interface of `intRo`, students are actually creating a fully-executable R script that they can download and run locally as well as viewing the script change real-time within the application. This code creation element of `intRo` is meant to generate excitement about programming in R and empower students to feel that they can generate code as well. `intRo` uses rmarkdown's render function in order to print the results, by dymanically executing the student's R script. By default, the output will include the R code, but if the student elects to hide the source code by clicking the code toggle button at the top, the code will not appear in the printed results.

This reproducibility framework has three key advantages. First, this eases a student who may be intimidated by programming into the idea that interacting with a user interface is really just a frontend for code. Seeing the correspondence between graphical clicks and printed code will hopefully also lessen the fear of coding that many students may have. Second, an analysis created by `intRo` can be reproduced in an R session to easily assess and extend the results. Finally, "printing" the results of an `intRo` analysis amounts to nothing more than executing the R code on the server, adding another layer of reproducibility.

@@

An `intRo` module is a set of self-contained executible R scripts that together produce a set of introductory statistics functionality. `intRo` modules were designed in this way to allow for simple dynamic creation of the user interface at run-time, as well as ease the process of converting existing analysis code to the `intRo` framework. A high-level diagram of this process is given in figure \ref{fig:app_creation_modules}. `intRo` modules are split up into multiple R scripts which are included either in Shiny's user interface or server definitions. At runtime, the `intRo` sources in the specified modules (contained in the modules folder) to dynamically generate the functionality available in the application. This allows for the specific functionality needed to be determined and adjusted by the individual course instructor. In this example, the instructor is electing to include a nonparametric module, which is not enabled by default, to allow the students to perform a wilcoxon rank sum test.

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/app_creation_modules.pdf}
\caption{This figure depicts how the Shiny \texttt{server.R} and \texttt{ui.R} files are populated using the modular structure within \texttt{intRo}. \texttt{intRo} modules are split up into multiple R scripts which are included either in Shiny's user interface or server definitions. At runtime, the \texttt{intRo} sources in the specified modules (contained in the modules folder) to dynamically generate the functionality available in the application. This allows for the specific functionality needed to be determined and adjusted by the individual course instructor. In this example, the instructor is electing to include a nonparametric module, which is not enabled by default, to allow the students to perform a wilcoxon rank sum test.}
\label{fig:app_creation_modules}
\end{figure*}

@@

Modularity was a design decision we focused on from the start of `intRo's` development. There are some practical benefits to thinking of related statistics and data science functionality in terms of modules. Course instructors can easily enable or disable functionality depending on the exact needs of the particular class being taught. Because this is done at run-time, including new modules is as simple as downloading and placing it within `intRo's` modules folder, or removing existing modules from that folder. Furthermore, errors can be more easily isolated to specific components. For instance, if an error is encountered, simply disabling the module can provide a temporary workaround while the issue is identified. Finally, modularity helps to organize the different pieces of code into functionality chunks that make it easier for developers to maintain.

@@

The modules provided with `intRo` are contained in the modules folder. The top level directory in the modules folder defines the category of the module (currently `data`, `summaries`, or `inference`). Within each of these categories is a folder named according to the name of the module. This folder houses the previously defined scripts. As an example, we will walk through the contents of the `nonparametric` module as previously mentioned in this section.

Since the `nonparametric` module performs a statistical test, it is part of the inference category, and hence can be found in the `intRo` repository at **modules/inference/nonparametric**. Let's first examine *helper.R*:

```{r, eval=FALSE}
nonparametrictest <- function(intro.data, x, y, 
                                conflevel, althyp, hypval) {
    interpolate(~(wilcox.test(x = df$x, y = df$y, 
                              conf.level = conf, 
                              alternative = althyp, 
                              mu = hypval)),
                  df = quote(intro.data),
                  x = x,
                  y = y,
                  conf = conflevel,
                  althyp = althyp,
                  hypval = hypval,
                  mydir = userdir, 
                  `_env` = environment(), 
                  file = "code_nonparametric.R")
}
```

This script is most immediately similar to standard R code. In this case, a function `nonparametrictest` is created which, depending on the values of the parameters, ultimately returns the result of a wilcoxon rank sum test. One important difference from a typical R script is that each call in the script is wrapped in a function called `interpolate` [@interpolate]. `interpolate` both executes the given R code on the server, and also writes the code executed to the script window at the bottom of `intRo`.

Because all the code needed to implement a wilcoxon rank sum test is found in the `base` and `stats` package, the `libraries.R` file is empty for the `nonparametric` module. Additionally, there are no reactive objects defined, so `reactive.R` is also empty. `observe.R`, which defines the Shiny observers needed, is reproduced below:

```{r, eval=FALSE}
observe({
    updateSelectizeInput(session, "group1_non", 
                         choices = intro.numericnames(), 
                         selected = ifelse(checkVariable(
                             intro.data(), input$group1_non), 
                             input$group1_non, 
                             intro.numericnames()[1]))
    updateSelectizeInput(session, "group2_non", 
                         choices = intro.numericnames(), 
                         selected = ifelse(checkVariable(
                             intro.data(), input$group2_non), 
                             input$group2_non, 
                             intro.numericnames()[2]))
})

observeEvent(input$store_nonparametric, {
    cat(paste0("\n\n", paste(readLines(
        file.path(userdir, "code_nonparametric.R")), 
        collapse = "\n")), 
        file = file.path(userdir, "code_All.R"), 
        append = TRUE)
})
```

Shiny observers are a class of reactive objects within the Shiny paradigm which do not return a value [@shiny]. For further discussion of reactivity, see section \ref{reactive-design-choices}. In this example, observers are created to ensure that the choices of variable for the `nonparametric` module are only numeric variables. This is accomplished by utilizing the global reactive `intro.numericnames()`, which returns a character vector containing the variables in the current dataset that are numeric. Finally, there is an event observer to store code generated from the module into the overall code script upon clicking the store button. The presence of this observer code and the definition of the button in the user interface are enforced, and must be present in any `intRo` module.

The `output.R` code is very simple:

```{r, eval=FALSE}
output$nonparametrictest <- renderPrint({
    return(nonparametrictable(intro.data(), input$group1_non, 
                              input$group2_non, input$conflevel_non, 
                              input$althyp_non, input$hypval_non))
})
```

The `output.R` script then simply uses Shiny's `renderPrint` function to display the resulting table.

@@

```{r, eval=FALSE}
nonparametric_ui <- tabPanel("Nonparametric",
    column(4,
         wellPanel(
             selectizeInput("group1_non", label = "Group 1 (x)", 
                            choices = numericNames(mpg), 
                            selected = numericNames(mpg)[1]),
             selectizeInput("group2_non", "Group 2 (y)", 
                            choices = numericNames(mpg), 
                            selected = numericNames(mpg)[2]),

@@

             selectizeInput("althyp_non", "Alternative Hypothesis", 
                            c("Two-Sided" = "two.sided", 
                              "Greater" = "greater", "Less" = "less")),
             numericInput("hypval_non", "Hypothesized Value",
                          value = 0),
             sliderInput("conflevel_non", "Confidence Level",
                         min=0.01, max=0.99, step=0.01, value=0.95),

             tags$button("", id = "store_nonparametric", type = "button", 
                         class = "btn action-button", list(icon("save"), 
                         "Store Nonparametric Result"), 
                         onclick = "$('#top-nav a:has(> .fa-print, 
                         .fa-code, .fa-download)').highlight();")
         )
    ),

    column(8,
         tags$b("Nonparametric Results"),
         verbatimTextOutput("nonparametrictest")
    )
)
```

@@

@@

`intRo` can be a powerful and effective tool for introductory statistics education. Its modular structure allows it to be flexible enough for many different applications and curriculums. Its ease-of-use allows the student to focus her attention on the statistics task at hand, rather than struggling with software licenses and confusing interface navigation. Reproducible code generated from each analysis can be used to spark an interest in R programming in those who might otherwise not be exposed to it.

@@

Challenges do exist with regards to the wider adoption of `intRo`. For instance, we will need to monitor how well the server hosting `intRo` handles the load of dozens of students performing data analyses at once. If performance issues are encountered, the infrastructure used may need to be expanded to handle current and future load. An unknown quantity will be how feasible it is to increase adoption of `intRo` across Iowa State, as well as to other universities. One limitation of `intRo` is that uploading a dataset beyond about 30,000 rows tends to be slow. Even once the data is successfully uploaded, the default modules produce results more slowly than with smaller datasets. This is a limitation that should be further investigated if and when `intRo` sees wider adoption.
